---
title: "Evaluating Cards in MTG Drafts"
description: |
  How good are we at figuring out which cards win?
author:
  - name: andrewstiegler
date: 2022-09-07
preview: preview_plot.png
output:
  distill::distill_article:
    self_contained: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
Drafting Magic: The Gathering (or anything, really) requires you to predict in advance the best picks for what you need. It's not easy to know in advance if you've made the right pick or not. Thanks to the team over at [17lands](https://www.17lands.com/), we can use the aggregated data of everyone who has installed their tracker to rate MTG cards. 
One thing that's nice about this data is that there is data for pick priority, as well as in-game performance for each card. The simplest ways of thinking about this are using the two metrics of "average taken at" (ATA), and "game-in-hand win-rate" (GIH WR). These metrics describe when a card is picked on average, and how often a card wins on average when it's actually in hand at some point in the game.
Here's a plot of ATA against GIH WR, using the latest available draft set, Dominaria United:

```{r}
library(jsonlite) #for parsing JSON data from 17lands and scryfall
library(tidyverse) #for data wrangling and cleaning
library(data.table) #for rbindlist later
library(plotly) #for interactive plots

#This function creates a well-formatted string for querying 17lands API
create_url_string <- function(set, start_date = "2019_01_01",
                              end_date = "2022_12_31"){
  
  end_date <- Sys.Date() %>% toString() %>% str_replace_all("-", "_")
  
  string <- paste("https://www.17lands.com/card_ratings/data?expansion=",
                  set,
                  "&format=PremierDraft",
                  "&start_date=",
                  start_date,
                  "&end_date=",
                  end_date,
                  sep = "")
  return(string)
}
#This function returns a character vector of each type extracted from 
#scryfall typeline
type_match <- function(string){
  types <- c("Sorcery", "Instant", "Creature", "Enchantment", "Artifact", "Land",
             "Equipment", "Aura", "Legendary", "Planeswalker")
  
  empty_list <- list()
  for(i in 1:length(types)){
    empty_list[[i]] <- as.numeric(grep(types[i], string))
    i <- i + 1
    
    real_type <- types[grep(1, empty_list)]
  }
  return(real_type)
}

# This function encodes rarity into a numeric
rarity_encoder <- function(string){
  if(string == "common"){rarity_enc <- 1}
  if(string == "uncommon"){rarity_enc <- 2}
  if(string == "rare"){rarity_enc <- 3}
  if(string == "mythic"){rarity_enc <- 4}
  
  return(rarity_enc)
}

# This function encodes color into a numeric (1-5 for monocolor, 
# 6 for "not monocolor")
color_encoder <- function(string){
  color_enc <- NA
  if(string == "W"){color_enc <- 1}
  if(string == "U"){color_enc <- 2}
  if(string == "R"){color_enc <- 3}
  if(string == "B"){color_enc <- 4}
  if(string == "G"){color_enc <- 5}
  if(string != "W" & string != "U" & string != "R" & string != "B" & string != "G"){color_enc <- 6}
  if(string == ""){color_enc <- 0}
  return(color_enc)
}

# This function downloads single-card data from scryfall based on url
extract_scryfall_per_card <- function(url){
  card_data_list <- fromJSON(paste("https://api.scryfall.com/cards/",
                                   url %>% substr(., 67, 102), sep = ""))
  
  type_string <- type_match(card_data_list$type_line)
  
  dual_face_check <- if("card_faces" %in% names(card_data_list)){1}else{0}
  
  card_tibble <- tibble("name" = card_data_list$name,
                        "colors" = toString(card_data_list$color_identity),
                        "cmc" = card_data_list$cmc,
                        "is_artifact" = if(sum(grep("Artifact", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_creature" = if(sum(grep("Creature", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_instant" = if(sum(grep("Instant", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_sorcery" = if(sum(grep("Sorcery", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_enchantment" = if(sum(grep("Enchantment", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_land" = if(sum(grep("Land", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_equipment" = if(sum(grep("Equipment", 
                                                    type_match(type_string))) > 0){1}else{0},
                        "is_planeswalker" = if(sum(grep("Planeswalker", 
                                                     type_match(type_string))) > 0){1}else{0},
                        "power" = if(sum(grep("Creature", 
                                              type_match(type_string))) > 0){card_data_list$power}else{0},
                        "toughness" = if(sum(grep("Creature", 
                                                  type_match(type_string))) > 0){card_data_list$toughness}else{0},
                        "text" = if(dual_face_check == 0){card_data_list$oracle_text}else{card_data_list$card_faces[1,5]},
                        "rarity" = rarity_encoder(card_data_list$rarity),
                        "pips" = str_count(card_data_list$mana_cost, "\\}"),
                        "url" = url)
  if(dual_face_check == 1){
    
    card_back_type_string <- type_match(card_data_list$card_faces[2,4])
    
    card_tibble$power <- if(sum(grep("Creature", 
                                     type_match(card_back_type_string))) > 0){card_data_list$card_faces[1,7]}else{0}
    card_tibble$toughness <- if(sum(grep("Creature", 
                                         type_match(card_back_type_string))) > 0){card_data_list$card_faces[1,8]}else{0}
    card_tibble$text <- card_data_list$card_faces[1,5]
    
    card_tibble_back <- tibble("name" = card_data_list$card_faces[2,2],
                               "colors" = toString(card_data_list$color_identity),
                               "cmc" = card_data_list$cmc,
                               "is_artifact" = if(sum(grep("Artifact", 
                                                           type_match(card_back_type_string))) > 0){1}else{0},
                               "is_creature" = if(sum(grep("Creature", 
                                                           type_match(card_back_type_string))) > 0){1}else{0},
                               "is_instant" = if(sum(grep("Instant", 
                                                          type_match(card_back_type_string))) > 0){1}else{0},
                               "is_sorcery" = if(sum(grep("Sorcery", 
                                                          type_match(card_back_type_string))) > 0){1}else{0},
                               "is_enchantment" = if(sum(grep("Enchantment", 
                                                              type_match(card_back_type_string))) > 0){1}else{0},
                               "is_land" = if(sum(grep("Land", 
                                                       type_match(card_back_type_string))) > 0){1}else{0},
                               "is_equipment" = if(sum(grep("Equipment", 
                                                            type_match(card_back_type_string))) > 0){1}else{0},
                               "is_planeswalker" = if(sum(grep("Planeswalker", 
                                                               type_match(card_back_type_string))) > 0){1}else{0},
                               "power" = if(sum(grep("Creature", 
                                                     type_match(card_back_type_string))) > 0){card_data_list$card_faces[2,7]}else{0},
                               "toughness" = if(sum(grep("Creature", 
                                                         type_match(card_back_type_string))) > 0){card_data_list$card_faces[2,8]}else{0},
                               "text" = card_data_list$card_faces[2,5],
                               "rarity" = rarity_encoder(card_data_list$rarity),
                               "url" = url)
    card_tibble <- rbind(card_tibble, card_tibble_back) %>% tibble() %>% relocate(url, .after = text)
    card_tibble$power <- as.numeric(card_tibble$power)
    card_tibble$toughness <- as.numeric(card_tibble$toughness)
    card_tibble <- card_tibble %>% select(name, colors, cmc, is_artifact, is_creature, is_instant,
                                          is_sorcery, is_enchantment, is_land, is_equipment,
                                          is_planeswalker, power, toughness, text, rarity, url)
  }
  
  card_tibble$color_enc <- sapply(card_tibble$colors, color_encoder)
  return(card_tibble)
}

#This function pulls in all set data from scryfall based on a processed 17lands dataset
extract_set <- function(data){
  urls <- data$url
  
  data_extracted <- list()
  for(i in 1:length(urls)){
    data_extracted[[i]] <- extract_scryfall_per_card(urls[i])
    Sys.sleep(0.15) #wait for the API
    i <- i + 1
  }
  data_tibble <- rbindlist(data_extracted, use.names = TRUE,
                           fill = TRUE) %>% tibble()
  
  return(data_tibble)
}

# This function pulls a 17lands/scryfall combo dataset when passed 3-letter set string
pull_set_data <- function(set_string){
  set_data <- fromJSON(create_url_string(set_string))
  
  # Extract data from scryfall for card info
  set_extracted <- extract_set(set_data)
  
  # Join 17lands and scryfall data
  set_joined <- full_join(set_extracted, set_data, by = "url")
  
  output_list <- list("17lands" = set_data,
                      "scryfall" = set_extracted,
                      "joined" = set_joined)
  return(output_list)
}

dmu_stats <- pull_set_data("DMU")

mtg_colors_full <- c("1" = "#F9FAF4", "2" = "#0E68AB", "3" = "#D3202A", 
                     "4" = "#150B01", "5" = "#01733E", 
                     "6" = "#D4AF37", 
                     "0" = "#c0c0c0")
mtg_colors_simple <- c("#c0c0c0", "#F9FAF4","#0E68AB","#D3202A", "#150B01","#01733E", 
                       "#D4AF37")

value_plot <- ggplot(dmu_stats[[3]])+
    geom_abline(slope = 0, intercept = 0.5)+
    geom_point(aes(x = avg_pick, y = ever_drawn_win_rate, 
                   text = name.x, fill = factor(color_enc)),
               shape = 21)+
    geom_smooth(aes(x = avg_pick, y = ever_drawn_win_rate))+
    scale_fill_manual(values = mtg_colors_simple)+
    xlab("Average Pick")+
    ylab("Overall Win Rate")+
    theme_classic()+
    theme(legend.position = "none")

ggsave("preview_plot.png", value_plot)
interactive_plot <- ggplotly(value_plot, tooltip = "text")
interactive_plot
```

You can mouse-over the plot and see the card names. Cards above the line are better than you'd expect from their pick rate. This makes sense with excellent cards (like [Archangel of Wrath](https://scryfall.com/card/dmu/3/archangel-of-wrath)). Cards below the line are worse than you'd expect (sorry [Urza Assembles the Titans](https://scryfall.com/card/dmu/37/urza-assembles-the-titans)).