---
title: "What's up with homeruns?"
description: |
  How have home run rates changed over time?
author:
  - name: andrewstiegler
date: 2022-09-01
output:
  distill::distill_article:
    self_contained: false
---
How have home run rates changed? Are there more homeruns now than ever?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
In order to look at baseball data in R, we need a good baseball database.
Luckily one exists for R already - the [Lahman dataset](https://www.seanlahman.com/baseball-archive/statistics/). 
In fact, there's even an R package available for it thanks to [Chris Dalzell and his team](https://github.com/cdalzell/Lahman). Let's use this dataset to figure out if homerun rates are high now compared to historical data. First, what's the average rate of home runs per plate appearance over time? Checking the data...  

```{r, fig.align = "center", out.width = "100%"}
library(Lahman)
library(tidyverse)
library(plotly)
library(tseries)

# Read data from Lahman's database of Batting
batting_df <- Batting

# Add function for getting player names
player_name_vec <- function(playerID){
  player_info <- playerInfo(playerID)
  player_name <- paste(player_info[2], player_info[3])
  return(player_name)
}

# Add PA stat and HR/PA
batting_data <- batting_df %>% mutate(PA = AB + BB + IBB + SH + SF)
batting_data <- batting_data %>% mutate(HR_rate = HR / PA)
batting_data <- batting_data %>% filter(yearID > 1955)
# batting_data$name <- sapply(batting_data$playerID, player_name_vec)

# Add the yearly average HR rate
yearly_rate <- batting_data %>% group_by(yearID) %>%
  summarise(HR_rate_mean = mean(HR_rate, na.rm = T))
colnames(yearly_rate) <- c("yearID", "HR_rate_mean")

batting_data <- full_join(batting_data, yearly_rate, by = "yearID")
batting_data$HR_normalized <- batting_data$HR_rate / batting_data$HR_rate_mean

hr_rate_over_time <- ggplot(batting_data, aes(x = yearID, y = HR_rate_mean))+
  geom_line()+
  geom_smooth()+
  xlab("Year")+
  ylab("HR per PA")+
  ggtitle("Homeruns per Plate Appearance, 1955-present")+
  theme_classic()

hr_rate_over_time_plotly <- ggplotly(hr_rate_over_time)
hr_rate_over_time_plotly
```


It looks like the rate increases over time, but it's noisy data. Let's check if the time series is stationary. We can use the Augmented Dickey-Fuller test
from the [tseries](https://cran.r-project.org/web/packages/tseries/index.html) package.

```{r}
adf_results <- yearly_rate %>% filter(yearID > 1955) %>% pull(HR_rate_mean) %>% adf.test()
adf_results
```
Based on this results (p-value > 0.05), we can conclude that there is a trend, and home run rates have been increasing overall in the past 70 years or so.

